---
title: 'PhD: CSE9099c - Weather Data Processing' 
author: "Jisha Joseph_1845"
date: "July-27-2018" 
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---
# Set up the environment
```{r}
rm(list=ls(all = T))
setwd("C:/Users/Hp/Desktop/INSOFE/Jisha Joseph_PhD")
```
# List & Load all required libraries here
```{r message = FALSE}
library(DMwR)
library(DataExplorer)
library(caret)
library(dplyr)
library(plyr)
library(ROCR)
library(corrplot)
library(class)
library(ada)
library(xgboost)
library(randomForest)
library(rpart)
library(C50)
library(e1071)
library(tidyverse)
library(lubridate)
library(tidyr)
library(sqldf)
library(ggplot2)
library(ROSE)
```

# Read & Understand the data
```{r}
train_data <- read.csv("Train.csv",header=TRUE)
test_data <- read.csv("Test.csv",header=TRUE)
```
## Keep an Original Copy of Train and test before doing any further processing
```{r}
train_data_orig<-train_data
test_data_orig <- test_data
```
## Get a feel of the data and get Insights
```{r}
head(train_data)
head(test_data)
tail(train_data)
tail(test_data)
```
## Structure & summary of the Train and Test data
```{r}
str(train_data)
summary(train_data)
summary(test_data) 
str(test_data)
```
# Target Attribute Derivation
## Convert ActualArrivalTimeStamp to POSIX type using Lubridate Library
```{r}
##Train data
train_data$ActualArrivalTS<-train_data[['ActualArrivalTimeStamp']]
train_data$ActualArrivalTS<-dmy_hm(train_data$ActualArrivalTS)
str(train_data)
head(train_data)
```
## Drop ActualArrivalTimeStamp as a new column has been created from it
```{r}
train_data$ActualArrivalTimeStamp=NULL
```
## Convert ScheduledArrTime and ScheduledDepTime to HH:MM format
```{r}
##Train Data
train_data$ScheduledArrTime<-sprintf("%04d",train_data$ScheduledArrTime)
format(strptime(train_data$ScheduledArrTime, format="%H%M"), format = "%H:%M")

train_data$ScheduledDepTime<-sprintf("%04d",train_data$ScheduledDepTime)
format(strptime(train_data$ScheduledDepTime, format="%H%M"), format = "%H:%M")

##Test Data

test_data$ScheduledArrTime<-sprintf("%04d",test_data$ScheduledArrTime)
format(strptime(test_data$ScheduledArrTime, format="%H%M"), format = "%H:%M")

test_data$ScheduledDepTime<-sprintf("%04d",test_data$ScheduledDepTime)
format(strptime(test_data$ScheduledDepTime, format="%H%M"), format = "%H:%M")
```
## Merge the Expected arrival/Departure into the same format of Actual Arrival/departure
```{r}

##Train Data
train_data<-unite(train_data,"ScheduledArrival",c("Year","Month","DayofMonth","ScheduledArrTime"),sep = '/',remove = FALSE)

train_data$ScheduledArrival<-ymd_hm(train_data$ScheduledArrival)
train_data$ScheduledArrTime=NULL

train_data<-unite(train_data,"ScheduledDep",c("Year","Month","DayofMonth","ScheduledDepTime"),sep = '/',remove = TRUE)
train_data$ScheduledDep<-ymd_hm(train_data$ScheduledDep)

##Test Data
test_data<-unite(test_data,"ScheduledArrival",c("Year","Month","DayofMonth","ScheduledArrTime"),sep = '/',remove = FALSE)
test_data$ScheduledArrival<-ymd_hm(test_data$ScheduledArrival)
test_data$ScheduledArrTime=NULL

test_data<-unite(test_data,"ScheduledDep",c("Year","Month","DayofMonth","ScheduledDepTime"),sep = '/',remove = TRUE)
test_data$ScheduledDep<-ymd_hm(test_data$ScheduledDep)
```
## Calculate the difference in Minutes
```{r}
difference <- difftime(train_data$ActualArrivalTS,train_data$ScheduledArrival, units='mins')
#difference
```
## Check if the difference is greater than 15 mins, then Create the Target attribute-FlightDelayStatus as 'Yes' else 'No'
```{r}
FlightDelayStatus<-ifelse(difference > 15,'Yes','No')
#FlightDelayStatus
```
## Add the FlightDelayStatus to the Train dataframe
```{r}
#train_data<-cbind(train_data,difference)
train_data<-cbind(train_data,FlightDelayStatus)
str(train_data)
str(test_data)
```
## Convert attributes to required data types
```{r}
## Train Data

train_data$DayOfWeek <- as.factor(as.character(train_data$DayOfWeek))

```
##Verify the structure of Train and test Data
```{r}
str(train_data)
str(test_data)
#There are extra levels in train for origin and destination compared to test
```
# Data exploration;
## Visualisations of Numeric attributes in Train Data
```{r}
library(purrr)
train_data_orig %>%
 keep(is.numeric) %>% 
 gather() %>% 
 ggplot(aes(value)) +
 facet_wrap(~ key, scales = "free") +
 geom_histogram()
```
#Check Uniquesness of Flightnumber
```{r}
unique(train_data$FlightNumber)
unique(test_data$FlightNumber)
```

## Univariate and Bivariate Analysis
* Univariate Analysis; Check if there are Outliers
```{r}
boxplot(train_data$ScheduledTravelTime)
```
* Bar Plot to check Kurtosis and Skewness
```{r}
ggplot(train_data,aes(x=FlightDelayStatus)) + geom_bar()
```
* Target is imbalanced as from the above plot.Hence have to balance the target
* Bivariate Analysis
```{r}
ggplot(train_data_orig,aes(x=Distance,
           y=ScheduledTravelTime)) + 
geom_point(size=2)
```

## Split Categorical and Numerical Attributes
```{r}
num_attr<-c("ScheduledDepTime","ScheduledArrTime","ScheduledTravelTime","Distance")
cat_attr<-setdiff(colnames(train_data),num_attr)
cat_attr
```
## Check Correlation Plot
```{r}
num_attr_df<-data.frame(train_data_orig[,num_attr])
#is.numeric((train_data_orig[,"FlightNumber"]))
cor_num_attr_df<-cor(num_attr_df)
#corrplot(cor_num_attr_df, method = "square")
corrplot(cor_num_attr_df, method = "number")
```
## Remove ScheduledTravelTime as it is highly correlated(0.98) with Distance
## Remove ScheduledDepTime as it is highly correlated with ScheduledArrTime
```{r}
## Keeping ScheduledDepTime and ScheduledArrTime for trial
train_data$ScheduledTravelTime=NULL
test_data$ScheduledTravelTime=NULL
```
# ###################Start of Weather Data Processing##########################
##Read AllStationsData_PHD as a dataframe
```{r}
AllStations<-read.table("AllStationsData_PHD.txt",sep="|", header=TRUE)
```
## Understand the data
```{r}
str(AllStations)
dim(AllStations)
```
## Split Categorical and Numerical Attributes for AllStations Data
```{r}
cat_attr_alls<-c("AirportID","TimeZone")
num_attr_alls<-setdiff(colnames(AllStations),cat_attr_alls)
num_attr_alls
```
## Check Correlation Plot
```{r}
num_attr_dfalls<-data.frame(AllStations[,num_attr_alls])
cor_num_attr_dfalls<-cor(num_attr_dfalls)
#corrplot(cor_num_attr_df, method = "square")
corrplot(cor_num_attr_dfalls, method = "number")
```
## Remove correlated attributes from AllStations Data before merging with Train Data
* Remove BarometerHeight as it is fully correlated with GroundHeight
* BarometerHeight and StationHeight are highly correlated.(0.84)
* Remove StationHeight as it is highly correlated with GroundHeight(0.84)
```{r}
AllStations$BarometerHeight=NULL
AllStations$StationHeight=NULL
```
## Missing Values Check and Imputation in AllStations
```{r}
sort(colSums(is.na(AllStations))/nrow(AllStations)*100, decreasing = T)
```
## Split Categorical and Numerical Attributes for AllStations Data after removal
```{r}
cat_attr_alls<-c("AirportID","TimeZone")
num_attr_alls<-setdiff(colnames(AllStations),cat_attr_alls)
num_attr_alls
```
## Chi-Square Test to identify correlated Categorical Attributes in AllStations
**p-value <0.05 pre-determined significant value. Hence the columns are highly dependent
```{r}
chisq.test(AllStations$AirportID, AllStations$TimeZone, correct=FALSE)
```
## Check Correlation Plot of updated AllStations Dataframe
```{r}
num_attr_dfalls<-data.frame(AllStations[,num_attr_alls])
cor_num_attr_dfalls<-cor(num_attr_dfalls)
#corrplot(cor_num_attr_df, method = "square")
corrplot(cor_num_attr_dfalls, method = "number")
```
## Check which all AirportIDs of Allstation data are present in Train Data Destination
```{r}
d<-unique(train_data$Destination[train_data$Destination %in% AllStations$AirportID])
o<-unique(train_data$Destination[train_data$Origin %in% AllStations$AirportID])
d
o
#Total 215 Unique AirportIDs(Origin and Destination)
```
## Check which all AirportIDs of Allstation data are present in Test Data Destination
```{r}
tsd<-unique(test_data$Destination[test_data$Destination %in% AllStations$AirportID])
tso<-unique(test_data$Destination[test_data$Origin %in% AllStations$AirportID])
tsd
tso
```
## Check for the Origin and Destination Levels mismatched in Train and test

```{r}
unique(train_data$Origin[!(train_data$Origin %in% test_data$Origin)])
```
## Use merge for joining AllStations and Train data,Test Data;
```{r}
train_allstde<-merge(x=train_data,y=AllStations,by.x=c("Destination"),by.y=c("AirportID"),all.x=TRUE)
train_allstor<-merge(x=train_data,y=AllStations,by.x=c("Origin"),by.y=c("AirportID"),all.x=TRUE)

train_allst<-rbind(train_allstor,train_allstde)


test_allstde<-merge(x=test_data,y=AllStations,by.x=c("Destination"),by.y=c("AirportID"),all.x=TRUE)
test_allstor<-merge(x=test_data,y=AllStations,by.x=c("Origin"),by.y=c("AirportID"),all.x=TRUE)
test_allst<-rbind(test_allstor,test_allstde)

rm(train_allstde,train_allstor,test_allstde,test_allstor)
```

#  Hourly and hpd processing
## Read 2004Hourly and HPD file to see the data
```{r}
df200401hrly<-read.table("200401hourly.txt",sep=",", header=TRUE)
df200401hpd<-read.table("200401hpd.txt",sep = ",",header = TRUE)
df200403hrly<-read.table("200403hourly.txt",sep=",", header=TRUE)
df200403hpd<-read.table("200403hpd.txt",sep = ",",header = TRUE)
df200405hrly<-read.table("200405hourly.txt",sep=",", header=TRUE)
df200405hpd<-read.table("200405hpd.txt",sep = ",",header = TRUE)
df200407hrly<-read.table("200407hourly.txt",sep=",", header=TRUE)
df200407hpd<-read.table("200407hpd.txt",sep = ",",header = TRUE)
df200409hrly<-read.table("200409hourly.txt",sep=",", header=TRUE)
df200409hpd<-read.table("200409hpd.txt",sep = ",",header = TRUE)
df200411hrly<-read.table("200411hourly.txt",sep=",", header=TRUE)
df200411hpd<-read.table("200411hpd.txt",sep = ",",header = TRUE)
```
## Merge all the hpd and hourly rows into one Dataframe
```{r}
df2004hrly<-rbind(df200401hrly,df200403hrly,df200405hrly,df200407hrly,df200409hrly,df200411hrly)
rm(df200403hrly,df200405hrly,df200407hrly,df200409hrly,df200411hrly)
df2004hpd<-rbind(df200401hpd,df200403hpd,df200405hpd,df200407hpd,df200409hpd,df200411hpd)
rm(df200403hpd,df200405hpd,df200407hpd,df200409hpd,df200411hpd)
#rm(df200401hpd) 
#rm(df200401hrly)
```
## Structure of HPD and Hourly
```{r}
str(df2004hpd)
str(df2004hrly)
```
## Missing Values Check and Imputation
```{r}
sort(colSums(is.na(df2004hrly))/nrow(df2004hrly)*100, decreasing = T)
sort(colSums(is.na(df2004hpd))/nrow(df2004hpd)*100, decreasing = T)
```
## Let's use Central imputation from DmWR package to impute missing values
```{r}
df2004hrly <- centralImputation(df2004hrly)
df2004hpd <- centralImputation(df2004hpd)
#df200401hpd<-centralImputation(df200401hpd) -->Testing Purpose
```
## Check if NA values are filled
```{r}
colSums(is.na(df2004hpd))
colSums(is.na(df2004hrly))
```

## Process hpd data 2004
## Convert and Merge the Timestamp in 2004hpd to Same as train_allst
## Convert Time to HH:MM format
```{r}
##HPD Train Data
# Convert Time to Numeric and then to HH:MM format
df2004hpd$Time<-as.numeric(df2004hpd$Time)
df2004hpd$Time<-sprintf("%04d",df2004hpd$Time)
format(strptime(df2004hpd$Time, format="%H%M"), format = "%H:%M")
str(df2004hpd)

###Testing Purpose
#df200401hpd$Time<-as.numeric(df200401hpd$Time)
#df200401hpd$Time<-sprintf("%04d",df200401hpd$Time)
#format(strptime(df200401hpd$Time, format="%H%M"), format = "%H:%M")
```

## Merge date and Time in hpd
```{r}
##hpd Train Data
df2004hpd<-unite(df2004hpd,"Timestamp",c("YearMonthDay","Time"),remove = TRUE)
df2004hpd$Timestamp<-ymd_hm(df2004hpd$Timestamp)
str(df2004hpd)
df2004hpdb<-df2004hpd

###Testing Purpose
#df200401hpd<-unite(df200401hpd,"Timestamp",c("YearMonthDay","Time"),remove = TRUE)
#df200401hpd$Timestamp<-ymd_hm(df200401hpd$Timestamp)
```

## Aggregate Time into 1 hour intervals for 2004hpd and Train data
```{r}
## Train Data
#############
# Aggregate each Timestamp into corresponding hour
# ScheduledArrival
traints<-cut.POSIXt(train_allst$ScheduledArrival, breaks = "hours")
traints
traindf<-as.data.frame.Date(traints)
str(traindf)
traindf$traints<-as.POSIXct(traindf$traints)
#traindf$traints<-traindf$traints + 3600
c<-cbind(train_allst,traindf)
train_allst<-c[,-3] #-->Remove ActualArrival Timestamp
colnames(train_allst)[14]<-"ExpectedArrival"

# ScheduledDep
traints<-cut.POSIXt(train_allst$ScheduledDep, breaks = "hours")
#traints
traindf<-as.data.frame.Date(traints)
str(traindf)
traindf$traints<-as.POSIXct(traindf$traints)
#traindf$traints<-traindf$traints + 7200
c<-cbind(train_allst,traindf)
train_allst<-c[,-3]  #-->Remove ActualDep Timestamp
colnames(train_allst)[14]<-"ExpectedDep"

# HPD data
##########
# Aggregate each Timestamp into corresponding hour
df2004hpdts<-cut.POSIXt(df2004hpd$Timestamp, breaks = "hours")
b<-as.data.frame.Date(df2004hpdts)
str(b)
c<-cbind(df2004hpd,b)
df2004hpd<-c[,-2]  ##-->Remove Actual Time
colnames(df2004hpd)[3]<-"Timestamp"


#Calculate the mean of HourlyPrecip grouped on WeatherStationID+Timestamp 
df2004hpd<-aggregate(HourlyPrecip~WeatherStationID+Timestamp,data=df2004hpd,FUN=function(df2004hpd) mean(df2004hpd))

###################################Trials###################################################
#df2004hpdagg<-aggregate(df2004hpd$WeatherStationID,list(hour=cut(as.POSIXct(df2004hp#d$Timestamp),"hour")),sum)

#df2004hpdxts <- xts(df2004hpd$WeatherStationID,as.POSIXct(df2004hpd$Timestamp)-1,df2#004hpd$HourlyPrecip)

#a<-aggregate(df200401hpdb[c("WeatherStationID","HourlyPrecip"),FUN=sum, #by=list(Timestamp=as.POSIXct(trunc(df200401hpdb$Timestamp, "hour"))))

                       
#a<-aggregate(df200401hpdb[c("HourlyPrecip")], FUN=, #by=list(d=as.POSIXct(trunc(df200401hpdb$Timestamp, "hour"))))
#a<-aggregate(. ~(Timestamp=as.POSIXct(trunc(df200401hpdb$Timestamp, "hour"))), #data=df200401hpdb, sum, na.rm=TRUE)                                                
#b<-list((Timestamp=as.POSIXct(trunc(df200401hpdb$Timestamp, #"hour"))),df200401hpdb$WeatherStationID,df200401hpdb$HourlyPrecip)
#c<-as.data.frame.Date(b)
#head(c)
#a<-aggregate((df200401hpdb[,1:2]), as.list(df200401hpdb[,3]), FUN = sum)

#a<-aggregate(df200401hpdb[c("HourlyPrecip")], FUN=mean, #by=list(d=as.POSIXct(trunc(df200401hpdb$Timestamp, "hour"))))

##Train Data
##a<-as.POSIXct(trunc(train_da$Timestamp, "hour"))
#b<-as.data.frame.Date(a)
#head(b)
#d<-cbind(df2004hpd,b)
#df2004hpd<-d[,-2]

#df2004hpd1<-sqldf("SELECT df2004hpd.* FROM df2004hpd GROUP BY #WeatherStationID,Timestamp")
# Add hour to the dataframe

## Group by
#library(plyr)
#groupColumns = c("WeatherStationID","a")
#dataColumns = c("HourlyPrecip")
#res = ddply(df2004hpd, groupColumns, function(x) mean(x[dataColumns]))
#res = ddply(df2004hpd, groupColumns, function(x) mean(x[dataColumns],na.rm = TRUE))
#?mean
#str(do.call(data.frame,aggregate(.~WeatherStationID+Timestamp,data=df200401hpd,FUN=f#unction(x) mean(x))))
############################################################################################
```
## Use merge for joining train_allst with hpd data
```{r}
trallst_hpdar<-merge(x=train_allst,y=df2004hpd,by.x=c("WeatherStationID","ExpectedArrival"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

trallst_hpd<-merge(x=trallst_hpdar,y=df2004hpd,by.x=c("WeatherStationID","ExpectedDep"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

trallsthpd<-trallst_hpd

trallst_hpd=transform(trallst_hpd,HourlyPrecip=((HourlyPrecip.x+HourlyPrecip.y)/2))
trallst_hpd<-trallst_hpd[,-c(15,16)]
str(trallst_hpd)
```

## Process hourly data 2004
## Split Categorical and Numerical Attributes for Hourly Data
```{r}
str(df2004hrly)
cat_attr_alls<-c("SkyConditions","Visibility","WindDirection","WindSpeed")
num_attr_alls<-setdiff(colnames(df2004hrly),cat_attr_alls)
num_attr_alls
```
## Chi-Square Test to identify correlated Categorical Attributes in hourly data
```{r}
df2004hrly$WindSpeed=as.factor(df2004hrly$WindSpeed)
chisq.test(df2004hrly$WindSpeed, df2004hrly$WindDirection, correct=FALSE)
chisq.test(df2004hrly$SkyConditions, df2004hrly$Visibility, correct=FALSE)
```
## Check Correlation Plot for df2004hrly
```{r}
num_attr_dfalls<-data.frame(df2004hrly[,num_attr_alls])
cor_num_attr_dfalls<-cor(num_attr_dfalls)
#corrplot(cor_num_attr_df, method = "square")
corrplot(cor_num_attr_dfalls, method = "number") 
```
## Remove correlated attributes
* Remove DewPointTemp as it is highly correlated(0.81) to DBT
* Remove WindDirection as it is highly correlated to Windspeed (p-value-2.2e-16)
* Remove Skyconditions as it is highly correlated to Visibility(p-value-2.2e-16)

```{r}
df2004hrly$DewPointTemp=NULL
df2004hrly$WindDirection=NULL
df2004hrly$SkyConditions=NULL
```

## Convert and Merge the Timestamp in 2004hourly to Same as trallst_hpd
## Convert ScheduledArrTime to HH:MM format
```{r}
##Hourly Train Data
# Convert Time to Numeric and then to HH:MM format
df2004hrlyb<-df2004hrly
df2004hrly$Time<-as.numeric(df2004hrly$Time)
df2004hrly$Time<-sprintf("%04d",df2004hrly$Time)
format(strptime(df2004hrly$Time, format="%H%M"), format = "%H:%M")
str(df2004hrly)

##Testing
#df200401hrlyb<-df200401hrly
#df200401hrly$Time<-as.numeric(df200401hrly$Time)
#df200401hrly$Time<-sprintf("%04d",df200401hrly$Time)
#format(strptime(df200401hrly$Time, format="%H%M"), format = "%H:%M")
#str(df200401hrly)
```

## Merge date and Time in hourly
```{r}
##hourly Train Data
df2004hrlyb<-df2004hrly
df2004hrly<-unite(df2004hrly,"Timestamp",c("YearMonthDay","Time"),remove = TRUE)
df2004hrly$Timestamp<-ymd_hm(df2004hrly$Timestamp)
str(df2004hrly)

###Testing Purpose
#df200401hrly<-unite(df200401hrly,"Timestamp",c("YearMonthDay","Time"),remove = TRUE)
#df200401hrly$Timestamp<-ymd_hm(df200401hrly$Timestamp)
```

## Aggregate Time into 1-hour-wide intervals for 2004hourly data
```{r}
# Hourly data
#############
# Aggregate each Timestamp into 1 hour window
df2004hrlyts<-cut.POSIXt(df2004hrly$Timestamp, breaks = "hours")
b<-as.data.frame.Date(df2004hrlyts)
str(b)
c<-cbind(df2004hrly,b)
df2004hrly<-c[,-2]
colnames(df2004hrly)[8]<-"Timestamp"

#Calculate the mean of HourlyPrecip grouped on WeatherStationID+Timestamp 
df2004hrly<-aggregate(cbind(Visibility,DBT,RelativeHumidityPercent,WindSpeed,WindGustValue,StationPressure)~WeatherStationID+Timestamp,data=df2004hrly,FUN=function(df2004hrly) mean(df2004hrly))

## Testing 01hourly
# Aggregate each Timestamp into 6 hour window

#df200401hrlyts<-cut.POSIXt(df200401hrly$Timestamp, breaks = "6 hours")
#b<-as.data.frame.Date(df200401hrlyts)
#str(b)
#c<-cbind(df200401hrly,b)
#df200401hrly<-c[,-2]
#colnames(df200401hrly)[8]<-"Timestamp"

#Calculate the mean of HourlyPrecip grouped on WeatherStationID+Timestamp 
#df200401hrly<-aggregate(cbind(Visibility,DBT,RelativeHumidityPercent,WindSpeed,WindG#ustValue,StationPressure)~WeatherStationID+Timestamp,data=df200401hrly,FUN=function(#df200401hrly) mean(df200401hrly))
```

## Merge hourly data with trallst_hpd
```{r}

trallshpd_hrar<-merge(x=trallst_hpd,y=df2004hrly,by.x=c("WeatherStationID","ExpectedArrival"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

trallsthpd_hrly<-merge(x=trallshpd_hrar,y=df2004hrly,by.x=c("WeatherStationID","ExpectedDep"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

trallsthpdhr<-trallsthpd_hrly
#trallsthpd_hrly<-trallsthpdhr

Visibility=rowMeans(trallsthpd_hrly[c('Visibility.x', 'Visibility.y')], na.rm=TRUE)
Visibility<-as.data.frame(Visibility)
trallsthpd_hrly<-cbind(trallsthpd_hrly,Visibility) #-->NaN Values

DBT=rowMeans(trallsthpd_hrly[c('DBT.x', 'DBT.y')], na.rm=TRUE)
DBT<-as.data.frame(DBT)
trallsthpd_hrly<-cbind(trallsthpd_hrly,DBT) #-->NaN Values

RelativeHumidityPercent=rowMeans(trallsthpd_hrly[c('RelativeHumidityPercent.x', 'RelativeHumidityPercent.y')], na.rm=TRUE)
RelativeHumidityPercent<-as.data.frame(RelativeHumidityPercent)
trallsthpd_hrly<-cbind(trallsthpd_hrly,RelativeHumidityPercent) #-->NaN Values

WindSpeed=rowMeans(trallsthpd_hrly[c('WindSpeed.x', 'WindSpeed.y')], na.rm=TRUE)
WindSpeed<-as.data.frame(WindSpeed)
trallsthpd_hrly<-cbind(trallsthpd_hrly,WindSpeed) #-->NaN Values

WindGustValue=rowMeans(trallsthpd_hrly[c('WindGustValue.x', 'WindGustValue.y')], na.rm=TRUE)
WindGustValue<-as.data.frame(WindGustValue)
trallsthpd_hrly<-cbind(trallsthpd_hrly,WindGustValue) #-->NaN Values

StationPressure=rowMeans(trallsthpd_hrly[c('StationPressure.x', 'StationPressure.y')], na.rm=TRUE)
StationPressure<-as.data.frame(StationPressure)
trallsthpd_hrly<-cbind(trallsthpd_hrly,StationPressure) #-->NaN Values
 
delete <- c('Visibility.x', 'Visibility.y','DBT.x', 'DBT.y','RelativeHumidityPercent.x', 'RelativeHumidityPercent.y','WindSpeed.x', 'WindSpeed.y','WindGustValue.x','WindGustValue.y','StationPressure.x', 'StationPressure.y')
trallsthpd_hrly<-trallsthpd_hrly[,!(colnames(trallsthpd_hrly) %in% delete),drop=FALSE]       

str(trallsthpd_hrly)
```

## Check for NA values in the final trallsthpd_hrly dataframe
## Missing Values Check and Imputation
```{r}
sort(colSums(is.na(trallsthpd_hrly))/nrow(trallsthpd_hrly)*100, decreasing = T)
```
## Let's use Central imputation from DmWR package to impute missing values
```{r}
trallsthpd_hrly <- centralImputation(trallsthpd_hrly)
```
## Check if NA values are filled
```{r}
colSums(is.na(trallsthpd_hrly))
trallsthpd_hrlybkp<-trallsthpd_hrly

```
## Aggregate the rows of trallsthpd_hrly into train rows 
```{r}
trallsthpd_hrly<-aggregate(cbind(DayOfWeek,Distance,FlightDelayStatus,GroundHeight,Longitude,HourlyPrecip,Visibility,DBT,RelativeHumidityPercent,WindSpeed,WindGustValue,StationPressure)~FlightNumber+ExpectedDep+ExpectedArrival,data=trallsthpd_hrly,FUN=function(trallsthpd_hrly) mean(trallsthpd_hrly))
trallsthpd_hrly$FlightDelayStatus <- ifelse(trallsthpd_hrly$FlightDelayStatus == '1', "No","Yes")
```
# Test hpd and hourly Processing
## Read 2005Hourly and HPD file to see the data
## Process 2005hpd data for test
```{r}
df200503hrly<-read.table("200503hourly.txt",sep=",", header=TRUE)
df200503hpd<-read.table("200503hpd.txt",sep = ",",header = TRUE)
df200507hrly<-read.table("200507hourly.txt",sep=",", header=TRUE)
df200507hpd<-read.table("200507hpd.txt",sep = ",",header = TRUE)
df200509hrly<-read.table("200509hourly.txt",sep=",", header=TRUE)
df200509hpd<-read.table("200509hpd.txt",sep = ",",header = TRUE)
df200511hrly<-read.table("200511hourly.txt",sep=",", header=TRUE)
df200511hpd<-read.table("200511hpd.txt",sep = ",",header = TRUE)
```
## Merge all the 2005hpd and hourly rows into one Dataframe for test
```{r}
df2005hrly<-rbind(df200503hrly,df200507hrly,df200509hrly,df200511hrly)
rm(df200503hrly,df200507hrly,df200509hrly,df200511hrly)
df2005hpd<-rbind(df200503hpd,df200507hpd,df200509hpd,df200511hpd)
rm(df200503hpd,df200507hpd,df200509hpd,df200511hpd)
```
## Structure of HPD and Hourly
```{r}
str(df2005hpd)
str(df2005hrly)
```
## Missing Values Check and Imputation
```{r}
sort(colSums(is.na(df2005hrly))/nrow(df2005hrly)*100, decreasing = T)
sort(colSums(is.na(df2005hpd))/nrow(df2005hpd)*100, decreasing = T)
```
## Let's use Central imputation from DmWR package to impute missing values
```{r}
df2005hrly <- centralImputation(df2005hrly)
df2005hpd <- centralImputation(df2005hpd)
```
## Check if NA values are filled
```{r}
colSums(is.na(df2005hpd))
colSums(is.na(df2005hrly))
```

## Process hpd data 2005
## Convert and Merge the Timestamp in 2005hpd to Same as train_allst
## Convert ScheduledArrTime to HH:MM format
```{r}
##HPD Test Data
# Convert Time to Numeric and then to HH:MM format
df2005hpd$Time<-as.numeric(df2005hpd$Time)
df2005hpd$Time<-sprintf("%04d",df2005hpd$Time)
format(strptime(df2005hpd$Time, format="%H%M"), format = "%H:%M")
str(df2005hpd)
```
## Merge date and Time in hpd
```{r}
##hpd Test Data
df2005hpd<-unite(df2005hpd,"Timestamp",c("YearMonthDay","Time"),remove = TRUE)
df2005hpd$Timestamp<-ymd_hm(df2005hpd$Timestamp)
str(df2005hpd)
df2005hpdb<-df2005hpd
```

## Aggregate Time into hour-wide intervals for 2005hpd and Test data
```{r}
## Test Data
#############
# Aggregate each Timestamp into corresponding hour

# Scheduled Arrival
str(test_allst)
tests<-cut.POSIXt(test_allst$ScheduledArrival, breaks = "hours")
tests
testdf<-as.data.frame.Date(tests)
str(testdf)
testdf$tests<-as.POSIXct(testdf$tests)
#testdf$tests<-testdf$tests + 3600
c<-cbind(test_allst,testdf)

test_allst<-c[,-3]  #-->Remove Actual Timestamp
colnames(test_allst)[12]<-"ExpectedArrival"

# ScheduledDeparture
tests<-cut.POSIXt(test_allst$ScheduledDep, breaks = "hours")
tests
testdf<-as.data.frame.Date(tests)
str(testdf)
testdf$tests<-as.POSIXct(testdf$tests)
testdf$tests<-testdf$tests + 3600
c<-cbind(test_allst,testdf)

test_allst<-c[,-3]  #-->Remove Actual Timestamp
colnames(test_allst)[12]<-"ExpectedDep"

# HPD data
##########
# Aggregate each Timestamp into corresponding hour
df2005hpdts<-cut.POSIXt(df2005hpd$Timestamp, breaks = "hours")
b<-as.data.frame.Date(df2005hpdts)
str(b)
c<-cbind(df2005hpd,b)
df2005hpd<-c[,-2]  ##-->Remove Actual Time
colnames(df2005hpd)[3]<-"Timestamp"

#Calculate the mean of HourlyPrecip grouped on WeatherStationID+Timestamp 
df2005hpd<-aggregate(HourlyPrecip~WeatherStationID+Timestamp,data=df2005hpd,FUN=function(df2005hpd) mean(df2005hpd))
```
## Use merge for joining test_allst with hpd data
```{r}

tsallst_hpdar<-merge(x=test_allst,y=df2005hpd,by.x=c("WeatherStationID","ExpectedArrival"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

tsallst_hpd<-merge(x=tsallst_hpdar,y=df2005hpd,by.x=c("WeatherStationID","ExpectedDep"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

tsallsthpd<-tsallst_hpd

tsallst_hpd=transform(tsallst_hpd,HourlyPrecip=((HourlyPrecip.x+HourlyPrecip.y)/2))

tsallst_hpd<-tsallst_hpd[,-c(13,14)] # Remove HourlyPrecip.x,y
str(tsallst_hpd)

```

## Remove correlated attributes as had removed in train due to high correlation
* Remove DewPointTemp as it is highly correlated(0.81) to DBT
* Remove WindDirection as it is highly correlated to Windspeed (p-value-2.2e-16)
* Remove Skyconditions as it is highly correlated to Visibility(p-value-2.2e-16)

```{r}
df2005hrly$DewPointTemp=NULL
df2005hrly$WindDirection=NULL
df2005hrly$SkyConditions=NULL
str(df2005hrly)
```

## Convert and Merge the Timestamp in 2005hourly to Same as tsallst_hpd
## Convert ScheduledArrTime to HH:MM format
```{r}
##Hourly Test Data
# Convert Time to Numeric and then to HH:MM format

df2005hrlyb<-df2005hrly
df2005hrly$Time<-as.numeric(df2005hrly$Time)
df2005hrly$Time<-sprintf("%04d",df2005hrly$Time)
format(strptime(df2005hrly$Time, format="%H%M"), format = "%H:%M")
str(df2005hrly)
```

## Merge date and Time in hourly test
```{r}
##hourly Test Data
df2005hrlyb<-df2005hrly
df2005hrly<-unite(df2005hrly,"Timestamp",c("YearMonthDay","Time"),remove = TRUE)
df2005hrly$Timestamp<-ymd_hm(df2005hrly$Timestamp)
str(df2005hrly)
```

## Aggregate Time into 1-hour-wide intervals for 2005hourly data
```{r}
# Hourly data
#############
# Aggregate each Timestamp into 1 hour window
df2005hrlyts<-cut.POSIXt(df2005hrly$Timestamp, breaks = "hours")
b<-as.data.frame.Date(df2005hrlyts)
str(b)
c<-cbind(df2005hrly,b)
df2005hrly<-c[,-2]
colnames(df2005hrly)[8]<-"Timestamp"

#Calculate the mean of all attributes in hourly data grouped on #WeatherStationID+Timestamp 
df2005hrly<-aggregate(cbind(Visibility,DBT,RelativeHumidityPercent,WindSpeed,WindGustValue,StationPressure)~WeatherStationID+Timestamp,data=df2005hrly,FUN=function(df2005hrly) mean(df2005hrly))
```

## Merge hourly data with tsallst_hpd
```{r}

tsallshpd_hrar<-merge(x=tsallst_hpd,y=df2005hrly,by.x=c("WeatherStationID","ExpectedArrival"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

tsallsthpd_hrly<-merge(x=tsallshpd_hrar,y=df2005hrly,by.x=c("WeatherStationID","ExpectedDep"),by.y=c("WeatherStationID","Timestamp"),all.x=TRUE)

tsallsthpdhr<-tsallsthpd_hrly
#tsallsthpd_hrly<-tsallsthpdhr

Visibility=rowMeans(tsallsthpd_hrly[c('Visibility.x', 'Visibility.y')], na.rm=TRUE)
Visibility<-as.data.frame(Visibility)
tsallsthpd_hrly<-cbind(tsallsthpd_hrly,Visibility) #-->NaN Values

DBT=rowMeans(tsallsthpd_hrly[c('DBT.x', 'DBT.y')], na.rm=TRUE)
DBT<-as.data.frame(DBT)
tsallsthpd_hrly<-cbind(tsallsthpd_hrly,DBT) #-->NaN Values

RelativeHumidityPercent=rowMeans(tsallsthpd_hrly[c('RelativeHumidityPercent.x', 'RelativeHumidityPercent.y')], na.rm=TRUE)
RelativeHumidityPercent<-as.data.frame(RelativeHumidityPercent)
tsallsthpd_hrly<-cbind(tsallsthpd_hrly,RelativeHumidityPercent) #-->NaN Values

WindSpeed=rowMeans(tsallsthpd_hrly[c('WindSpeed.x', 'WindSpeed.y')], na.rm=TRUE)
WindSpeed<-as.data.frame(WindSpeed)
tsallsthpd_hrly<-cbind(tsallsthpd_hrly,WindSpeed) #-->NaN Values

WindGustValue=rowMeans(tsallsthpd_hrly[c('WindGustValue.x', 'WindGustValue.y')], na.rm=TRUE)
WindGustValue<-as.data.frame(WindGustValue)
tsallsthpd_hrly<-cbind(tsallsthpd_hrly,WindGustValue) #-->NaN Values

StationPressure=rowMeans(tsallsthpd_hrly[c('StationPressure.x', 'StationPressure.y')], na.rm=TRUE)
StationPressure<-as.data.frame(StationPressure)
tsallsthpd_hrly<-cbind(tsallsthpd_hrly,StationPressure) #-->NaN Values
 
delete <- c('Visibility.x', 'Visibility.y','DBT.x', 'DBT.y','RelativeHumidityPercent.x', 'RelativeHumidityPercent.y','WindSpeed.x', 'WindSpeed.y','WindGustValue.x','WindGustValue.y','StationPressure.x', 'StationPressure.y')
tsallsthpd_hrly<-tsallsthpd_hrly[,!(colnames(tsallsthpd_hrly) %in% delete),drop=FALSE]       

str(tsallsthpd_hrly)
```
## Check for NA values in the final tsallsthpd_hrly dataframe
## Missing Values Check and Imputation
```{r}
sort(colSums(is.na(tsallsthpd_hrly))/nrow(tsallsthpd_hrly)*100, decreasing = T)
```
## Let's use Central imputation from DmWR package to impute missing values
```{r}
tsallsthpd_hrly <- centralImputation(tsallsthpd_hrly)
```
## Check if NA values are filled
```{r}
colSums(is.na(tsallsthpd_hrly))
```
## Aggregate the rows of tsallsthpd_hrly into test rows 
```{r}

tsallsthpd_hrly<-aggregate(cbind(DayOfWeek,Distance,GroundHeight,Longitude,HourlyPrecip,Visibility,DBT,RelativeHumidityPercent,WindSpeed,WindGustValue,StationPressure)~FlightNumber+ExpectedDep+ExpectedArrival,data=tsallsthpd_hrly,FUN=function(tsallsthpd_hrly) mean(tsallsthpd_hrly))

```
##   #######################End of Weather Data Processing#################
