---
title: 'PhD: CSE9099c - Machine Learning Algorithms with Weather Data-H2O'
author: "Jisha Joseph 1845"
date: "July-27-2018"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
---
# Set up the environement
```{r}
rm(list=ls(all = T))
setwd("C:/Users/Hp/Desktop/INSOFE/Jisha Joseph_PhD") 
```
# List & Load all required libraries here
```{r message = FALSE}
library(DMwR)
library(caret)
library(ROCR)
library(corrplot)
library(class)
library(ada)
library(xgboost)
library(randomForest)
library(rpart)
library(C50)
library(e1071)
library(tidyverse)
library(lubridate)
library(tidyr)
library(h2o)
library(ggplot2)
library(MLmetrics)
```
##   #######################End of Weather Data Processing#################
## Full Train Data Pre-processing
## Data exploration;
## Visualisations of Numeric attributes in Train Data
```{r}
library(purrr)
trallsthpd_hrly %>%
 keep(is.numeric) %>% 
 gather() %>% 
 ggplot(aes(value)) +
 facet_wrap(~ key, scales = "free") +
 geom_histogram()

```
## Check attributes of near zero variance in Train data
```{r}
#tr_bkp<-train_data
#ts_bkp<-test_data
train_data<-trallsthpd_hrly
test_data<-tsallsthpd_hrly
nearZeroVar(train_data, freqCut = 95/5, uniqueCut = 10, saveMetrics = FALSE,names = TRUE, foreach = FALSE, allowParallel = TRUE)
```
### Keeping HourlyPrecip and WindGustValue for testing purpose
```{r}
#delete <- c("HourlyPrecip")
#train_data<-train_data[,!(colnames(train_data) %in% delete),drop=FALSE]
#test_data<-test_data[,!(colnames(test_data) %in% delete),drop=FALSE]
str(train_data)
```
## Split Categorical and Numerical Attributes
```{r}
cat_attr<-c("FlightNumber","FlightDelayStatus","ExpectedArrival","ExpectedDep")
num_attr<-setdiff(colnames(train_data),cat_attr)
num_attr
```
## Check Correlation Plot
```{r}
num_attr_df<-data.frame(train_data[,num_attr])
cor_num_attr_df<-cor(num_attr_df)
#corrplot(cor_num_attr_df, method = "square")
corrplot(cor_num_attr_df, method = "number")
```

* Remove GroundHeight as it is highly negatively correlated to StationPressure(-0.96)
```{r}

delete <- c("GroundHeight")
train_data<-train_data[,!(colnames(train_data) %in% delete),drop=FALSE]
test_data<-test_data[,!(colnames(test_data) %in% delete),drop=FALSE]

colnames(train_data)
```
## Backups of Data and Removal of attributes as needed
```{r}
## Backups of main data
trn_data<-train_data
tst_data<-test_data

#train_data<-trn_data
#test_data<-tst_data

str(FlightNumber)
train_data$FlightDelayStatus<-as.factor(as.character(train_data$FlightDelayStatus))

FlightNumber<-as.data.frame(test_data_orig$FlightNumber)
colnames(FlightNumber)[1]<-"FlightDelayStatus"
test_data_trial<-test_data
test_data<-test_data[match(FlightNumber$FlightDelayStatus,test_data$FlightNumber),]
train_data$FlightNumber<-NULL
test_data$FlightNumber<-NULL

colnames(test_data)
```
## Bivariate Analysis
* DBT and RelativeHumidityPercent
```{r}
ggplot(train_data,aes(x=DBT,
           y=RelativeHumidityPercent)) + 
geom_point(size=2)+ggtitle("DBT VS RelativeHumidityPercent")
```
# ############################### Basic Model Building #############################
## Check for the proportion of Target distribution of classes
```{r}
table(train_data$FlightDelayStatus)
str(train_data)
```
* Confirm equal distribution of Status in train & validation.
```{r}
prop.table(table(train_data$FlightDelayStatus))

train_data$ExpectedArrival=as.numeric(train_data$ExpectedArrival)
test_data$ExpectedArrival=as.numeric(test_data$ExpectedArrival)
train_data$ExpectedDep=as.numeric(train_data$ExpectedDep)
test_data$ExpectedDep=as.numeric(test_data$ExpectedDep)
``` 

# ############################### Basic Model Building #############################
## Train-Validation split
```{r}
train_data_df<-train_data
#train_data<-train_data_df
set.seed(123)
train_rows <- createDataPartition(train_data$FlightDelayStatus, p = 0.80, list = F )
pre_train <- train_data[train_rows, ]
pre_val <- train_data[-train_rows, ]

validation_data <- train_data[-train_rows, ]
train_data<-train_data[train_rows, ]

```

## Check for the proportion of Target distribution of classes
```{r}
table(train_data$FlightDelayStatus)
table(validation_data$FlightDelayStatus)
```

## H2O
## H2O - Random Forest
```{r}
## Reordering Train and Validation data
#train_data<-train_data[c(1:4,6:13,5)]
#validation_data<-validation_data[c(1:4,6:13,5)]

localH2O <- h2o.init(nthreads = -1)

#data to h2o cluster
train_h2o <- as.h2o(train_data)
val_h2o<-as.h2o(validation_data)
test_h2o <- as.h2o(test_data)

#check column index number
colnames(train_h2o) 
y_dep <- 13

#Independent variables 
x_indep <- c(1:12)


#sample_factors <- c(.25,5.)

H2O_RF_Model = h2o.randomForest(y = y_dep, x = x_indep,training_frame = train_h2o,ntrees = 75,max_depth = 100,min_rows = 15,seed = 12345,mtries =7,balance_classes=TRUE)

#hyper_params <- list(class_sampling_factors = list(c(.25, 0.5),c(2.,3.)))

#H2O_RF_Model = h2o.randomForest(y = y_dep, x = x_indep,training_frame = train_h2o,nfolds = 9,ntrees = 75,max_depth = 40,min_rows = 30,seed = 12345,mtries =5,balance_classes=TRUE,calibrate_model = TRUE,calibration_frame = val_h2o,class_sampling_factors = sample_factors,binomial_double_trees=TRUE)
#H2O_RF_Model = h2o.grid(y = y_dep, x = x_indep,training_frame = train_h2o,nfolds = 7,ntrees = 85,max_depth = 7,min_rows = 20,seed = 12345,mtries =7,nfolds = 9,balance_classes=TRUE,calibrate_model = TRUE,grid_id = "covtype_grid",algorithm = "drf",calibration_frame = val_h2o,hyper_params = hyper_params,search_criteria = list(strategy = "Cartesian"))

# Sort the grid models by logloss
#sortedGrid <- h2o.getGrid("covtype_grid", sort_by = "logloss", decreasing = FALSE)
#sortedGrid
#mtries = 4 , Let it use the default sqrt(p) for classification

```
## Check the performance of model
```{r}
h2o.performance(H2O_RF_Model)
```
## Check for important variables
```{r}
h2o.varimp(H2O_RF_Model)
```
## Plot the Variable Importance for TOP 10 features
```{r}
h2o.varimp_plot(H2O_RF_Model)
```
## Make prediction on train data using the model
```{r}
Pred_Train_H2O_RF = as.data.frame(h2o.predict(H2O_RF_Model , train_h2o))
```
## Create a confusion Matrix for Train
```{r}
#Pred_Train_RF <- ifelse(Pred_Train_H2O_RF$No < 0.5, "Yes","No")
Pred_Train_RF<-Pred_Train_H2O_RF$predict
conf_matrix <- table(train_data$FlightDelayStatus, Pred_Train_RF)
print(conf_matrix)
```
## Make prediction on Validation data using the model
```{r}
Pred_Validate_H2O_RF = as.data.frame(h2o.predict(H2O_RF_Model , val_h2o))
```
## Create a confusion Matrix for Validation -->F1 Score-->86.20
```{r}
#Pred_val_RF <- ifelse(Pred_Validate_H2O_RF$No <= 0.5, "Yes","No")
Pred_val_RF <- Pred_Validate_H2O_RF$predict
conf_matrix <- table(validation_data$FlightDelayStatus, Pred_val_RF)
print(conf_matrix)

F1_Score(validation_data$FlightDelayStatus, Pred_val_RF)
```
## Make prediction on Validation data using the model
```{r}
Pred_Test_H2O_RF = as.data.frame(h2o.predict(H2O_RF_Model , test_h2o))
```
## Prediction F1 Score-->28.13
```{r}
#FlightDelayStatus <- ifelse(Pred_Test_H2O_RF$No <= 0.6, "Yes","No")
FlightDelayStatus<-Pred_Test_H2O_RF$predict
FlightNumber<-as.data.frame(FlightNumber)
preds_h2o_rf<-cbind(FlightNumber,FlightDelayStatus)
write.csv(preds_h2o_rf,"submission_h2Orf.csv",row.names = F)
```
## H2O
## H2O - Gradient Boosting
```{r}
## Reordering Train and Validation data
#train_data<-train_data[c(1:4,6:13,5)]
#validation_data<-validation_data[c(1:4,6:13,5)]

localH2O <- h2o.init(nthreads = -1)

#data to h2o cluster
train_h2o <- as.h2o(train_data)
val_h2o<-as.h2o(validation_data)
test_h2o <- as.h2o(test_data)

#check column index number
colnames(train_h2o) 
y_dep <- 13

#Independent variables    
x_indep <- c(1:12)

H2O_gbm_Model = h2o.gbm(y = y_dep, x = x_indep,training_frame = train_h2o,ntrees = 500,nfolds = 7,max_depth = 215,min_rows = 5,seed = 12345,learn_rate = 0.05,learn_rate_annealing=0.90,distribution ="bernoulli",sample_rate = .05,col_sample_rate=0.3,balance_classes=TRUE)

```
## Check the performance of model
```{r}
h2o.performance(H2O_gbm_Model)
```
## Check for important variables
```{r}
h2o.varimp(H2O_gbm_Model)
```
## Plot the Variable Importance for TOP 10 features
```{r}
h2o.varimp_plot(H2O_gbm_Model)
```
## Make prediction on train data using the model
```{r}
Pred_Train_H2O_gbm = as.data.frame(h2o.predict(H2O_gbm_Model , train_h2o))
```
## Create a confusion Matrix for Train
```{r}
Pred_Train_gbm<-Pred_Train_H2O_gbm$predict
conf_matrix <- table(train_data$FlightDelayStatus, Pred_Train_gbm)
print(conf_matrix)
```
## Make prediction on Validation data using the model
```{r}
Pred_Validate_H2O_gbm = as.data.frame(h2o.predict(H2O_gbm_Model , val_h2o))
```
## Create a confusion Matrix for Validation -->F1 Score-->80.73
```{r}
Pred_val_gbm <- Pred_Validate_H2O_gbm$predict
conf_matrix <- table(validation_data$FlightDelayStatus, Pred_val_gbm)
print(conf_matrix)

F1_Score(validation_data$FlightDelayStatus, Pred_val_gbm)
```
## Make prediction on Validation data using the model
```{r}
Pred_Test_H2O_gbm = as.data.frame(h2o.predict(H2O_gbm_Model , test_h2o))
```
## Prediction F1 Score-->28.13
```{r}
FlightDelayStatus<-Pred_Test_H2O_gbm$predict
FlightNumber<-as.data.frame(FlightNumber)
preds_h2o_gbm<-cbind(FlightNumber,FlightDelayStatus)
write.csv(preds_h2o_gbm,"submission_h2Ogbm.csv",row.names = F)
```